---
title: "Telling stories with charts - visualizations with ggplot2"
author: "Adi Sarid"
output: html_document
---

# What for and why now?

When we think about the data science workflow, it makes a lot of sense to start our lesson with tidying data or data transformations, however, I chose to start with visualizations. Why is that?

I'm adopting the approach of the "R for Data Science book", which talks about visualizations in one of its early chapters (even before data transformations), because charts give a lot of motivation to our next chapters and also helps us develop a way of thinking.

As you will see, the following exercises will also contain some elements of tidying and transforming data (because sometimes its a necessity towards working on a chart). In a future lesson we will delve deeper and extend our knowledge into tidying and data transformations.

**Remember:**
When you do the exercises, if you get stuck, put on a pink sticky note. If I haven't noticed it, call me. 

Along the exercise we will have "checkpoints", if you reached a checkpoint put on your green sticky note. You can take a break, dring a soda, or continue to the next exercise. Once I see a lot of green sticky notes we will stop and discuss the exercise up to the checkpoint.

# Exercise 1: the *google play* data set

In this exercise we will work with a file downloaded from the competition website kaggle. You can see the details [here](https://www.kaggle.com/lava18/google-play-store-apps). You are going to load the file directly from our course's repository, using the following command:

```{r load googleplay scraped data file}

suppressMessages(
   library(tidyverse)
   )
google_play_raw <- read_csv(file = "https://raw.githubusercontent.com/adisarid/Riskified_training/master/datasets/googleplaystore.csv")

```

**Question 1:** Did you notice I used the `supressMessages(...)` function. What does it do? Why did I use it?

**Question 2:** Did you notice any "parsing failures" when reading the file? 
Try to figure out what do they mean, and think about how would you solve the problem. The error message contains a row location so you can look at the data in the area of the error to try and figure it out, e.g.: 

```
location <- ???
google_play_raw[(location-3):(location+3),]
```

Once you figure out what's wrong, filter this line out of the file (but note that there are other ways to handle this).

```
google_play_raw <- google_play_raw[-location,]
```

**Question 3**: Is there any other way you could have filtered this messy line? (think about `filter()`, the `is.na()` and `!` operator)

Look at the dataset using `glimpse(???)`, notice that most columns were read as characters and a few as double.

**Question 4:** What variables were read as character but you think that you would be better off if they were tranformed to a different type?

The `Category` variable represents the app's category. 

**Question 5:** Complete the code below and answer the questions that follow. Note that the last part of the code (`theme(...)`) is meant to rotate the x axis labels, for better readability. First try without it and then add it.

```
ggplot(data = ???, mapping = aes(x = Category)) + 
   geom_bar() + 
   theme(axis.???.??? = element_text(angle = ???))
```


   1. What is the order by which the columns are sorted?

Build a chart that shows how many apps are supported by each version (android version on the x-axis and number of apps on the y-axis). Note that in the code we used ` which is required when the variable name contains non-standard characters (e.g. spaces or Hebrew).

```
ggplot(???, aes(x = `Android Ver`)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 45))
```

**Question 6:** 

   1. What was the latest android version when the file was generated?
   2. Would you say that some categories of `Android Ver` should be grouped together?

In the following segment we will examine the relationship between the rating of an app and the number of users who rated it. Build a chart that will help you examine this relationship.

Hint: use a `geom_point()` with x as number of reviews and y as the average rank.

```
ggplot(XXX, aes(XXX, XXX)) + 
   geom_XXX() + 
   stat_smooth(method = "XXX")
```

**Question 7:** 

   1. In the documentation of `stat_smooth` there are a number of smoothing methods. What did you choose and why?
   2. Do you see anything wrong with the chart?
   3. Bonus: is there a transformation (**mathematical function**) which you can use on the x-axis, for the chart to make more sense? What is it? What do you see when you use it, that you didn't see before?
   
**Question 8:** Use the function `cut` to split the number of reviews to five different groups. Generate a chart of rank versus reviews in which each new reviews-group is colored in different color.

```
# note the use of tidyverse here (%>%, mutate, etc.)
google_play_groups <- google_play %>% 
   mutate(reviews_group = cut(XXX, breaks = c(10^(0:6), max(Reviews) + 1)))
   
rating_reviews_grouped_chart <- 
  ggplot(google_play_groups, aes(x = XXX, y = XXX, color = XXX)) + 
  geom_point() + 
  stat_smooth(method = "lm") +
  scale_x_log10()
```

   1. Do you identify any new relationships?
   2. Do specific ranges have stronger relationships than other ranges?

**Question 9:** Use a boxplot to compare the distribution of the number of reviews between free and paid apps

   1. Do you see a redundent category on the x-axis (NaN)? try to get rid of it using `filter()` and `is.nan()`. Did it work? how would you get rid of it?
   2. Did you notice the error message about infinite values when you plotted the data? what does the error message mean?
   2. What does this comparison between paid and free tells you?
   3. Use a `log()` transformation on the number of reviews, and re-plot. What do you see now?
   4. Describe the difference between using `log(Reviews)` versus `scale_y_log()`.
   5. Compare the distribution of average rank of free apps versus paid apps. What does this comparison tells you?

**Question 10:** An alternative way to examine distributions is with a histogram. In this question we will examine the distribution of app size.

   1. What is the problem with the app size variable?
   2. Use `mutate`, `str_replace` (from package `stringr`) and the function `as.numeric` to convert it.
   3. Use the functions `geom_histogram` and `geom_freqpoly`. Both has a parameter called `bins`. It's default value is 30. Try to examine different values. What does it affect? What happens when you choose the same value for both? what happens when you choose different values?
   4. Where is the "main mass" of the distribution?
   
```
google_play <- google_play %>%
   mutate(size_app_numeric = as.numeric(str_replace(XXX, "M", ""))
   
ggplot(XXX, aes(XXX)) + 
   geom_histogram(fill = "lightblue", bins = XXX) + 
   geom_freqpoly(size = 1.5, bins = XXX)
```

**Question 11:** One capability of `ggplot2` is splitting the charts with facets. To the chart you used of average rating versus log(reviews), add a facet (`facet_wrap`) by the variable Category, i.e., add `+ facet_wrap(~ ???)`.

   1. What happened to the trend you identified earlier?
   2. Compare the EDUCATION category versus TOOLS. What differences do you see and what is their meaning?

**Checkpoint 1: If you reached here you have completed the first exercise. Hurray!.**

   a. Stick your green sticky note. 
   b. Come to me and receive your brand new `ggplot2` badge. 
   c. You can take a break and/or continue with the exercises. 
   d. When I see enough green sticky notes around we will solve the last exercise together.

## Exercise 1.5: A very useful trick - automated parameterized reports (with RMarkdown)

This trick is not directly related to plotting, but it seems appropriete to show it now, after we have done a few ggplots and now want to automate our process. 

Suppose that there are difference moderators that are in charge of each "google play category". You want to provide each moderator with its own data but not others'. This means generating 33 RMarkdown reports (the number of categories), or does it?

Open file `02-Automated-report-sample.Rmd`. For now, skip the first bit that is in the ---. It is called yaml (these are instructions to the RMarkdown compiler).Review the code, mainly it is taken from the exercise you have just solved.

Knit the RMarkdown into an html.

Now, do the same but instead of clicking on knit, open the knit menu (small arrow next to the knit icon), and click *knit with parameters*. Examin what options you have and look back at the *yaml* to see how they correspond to the *yaml* settings. If this didn't work, you might need to install `shiny` using: `install.packages("shiny")`.

To automate the knitting process you can use the following command (try to run it for category "BUSINESS" with app size > 50):

```
rmarkdown::render(input = ???,
                  params = list(??? = ???, ??? = ???),
                  output_file = ???,
                  output_dir = ???)
```

When we will learn the `purrr` package, you will see how you can create a script that will render all the categories there are with just a few lines of code.

We now go back to flow of plotting and discuss how ggplot2 protects you against shooting yourself in the leg.

**Checkpoint 2: If you reached here you have completed the 1.5 exercise. Awesome.**

   a. Stick your green sticky note. 
   b. You can take a break and/or continue with the exercises. 
   c. When I see enough green sticky notes around we will solve the last exercise together.

# Exercise 2: Some graphing pifalls and how ggplot2 protects you against them

Some of the following was inspired by this [blog post](https://blog.funnel.io/why-we-dont-use-pie-charts-and-some-tips-on-better-data-visualizations), and by Maor Shoshan which emailed me a couple of requests (**and I encourage all of you to do the same!**).

## The double y-axis paradigm 

Double y-axis can be useful sometimes but can also be misleading. `ggplot2` makes it "hard" to generate a double y-axis, with a sound philosophy behind. See [Hadley's response](https://stackoverflow.com/questions/3099219/plot-with-2-y-axes-one-y-axis-on-the-left-and-another-y-axis-on-the-right):

   It's not possible in `ggplot2` because I believe plots with separate y scales (not y-scales that are transofrmations of each other) are fundamentally flawed...
   
Let's try to analyze this. 

**Question 1:** In the following example, what is the relationship between Bannana prices and Computer prices? what is the meaning of their intersection around the second month?

```{r double y-axis}

sales <- tribble(
   ~Month, ~Bannana, ~Computer,
   1, 5.3, 3000,
   2, 5.5, 2950,
   3, 6, 2500,
   4, 5.9, 2600,
   5, 5.9, 2450,
   6, 5.6, 2320,
   7, 5.7, 2400
) %>% 
   gather(item, price, -Month) %>% 
   mutate(price = ifelse(item == "Computer", price/3000*6, price))

ggplot(sales, aes(x = Month, y = price, color = item)) + 
   geom_line(size = 2) + 
   scale_y_continuous("Bannana prices", sec.axis = sec_axis(~ .*3000/6, name = "Computer prices")) + 
   theme_minimal() + 
   theme(legend.position = "top") + 
   ggtitle("A misleading double y-axis")
```

**Question 2:** Now examine the following chart. Non of the numbers or values changed, just the transformation that positions the computer's axis. The two lines seem to coinside more, but what does that mean?

```{r second alternative}
sales <- tribble(
   ~Month, ~Bannana, ~Computer,
   1, 5.3, 3000,
   2, 5.5, 2950,
   3, 6, 2500,
   4, 5.9, 2600,
   5, 5.9, 2450,
   6, 5.6, 2320,
   7, 5.7, 2400
) %>% 
   gather(item, price, -Month) %>% 
   mutate(price = ifelse(item == "Computer", price/2500*6, price))

ggplot(sales, aes(x = Month, y = price, color = item)) + 
   geom_line(size = 2) + 
   scale_y_continuous("Bannana prices", sec.axis = sec_axis(~ .*3000/6, name = "Computer prices")) + 
   theme_minimal() + 
   theme(legend.position = "top") + 
   ggtitle("A misleading double y-axis: #2")
```

**Question 3:** There are a few cases in which a double y-axis is OK (and may even contribute).

   1. When would using a dual y-axis be beneficial?
   2. Download (or make up) relevant data and illustrate it via a `ggplot2` chart with the `sec.axis` function/argument.

## Pie charts

The problem with pie charts is that they require much more effort. They render one-dimentional data on a two dimentional plane, so in fact pie charts **do not exist in `ggplot2`!** (there are some blog posts talking about how to make a pie chart with `coord_polar`, and they are really long).

Base-r though has pie charts. DO NOT TRY THIS AT HOME:

```{r pie versus bar}
fruits <- tribble(
   ~fruit, ~portion,
   "Bannanas", 0.33,
   "Apples", 0.29,
   "Strawberries", 0.19,
   "Pares", 0.1,
   "Blueberries", 0.07,
   "Oranges", 0.02
)

# DO NOT TRY THIS AT HOME
pie(fruits$portion, labels = fruits$fruit, main = "DO NOT TRY THIS AT HOME!!!")
```

**Question 1:** How would you plot the `fruits` data in `ggplot2`'s ecosystem? (think bars) and do that.

**Checkpoint 3: If you reached here you have completed exercise 2. Great job!.**

   a. Stick your green sticky note. 
   b. You can take a break and/or continue with the exercises. 
   c. When I see enough green sticky notes around we will solve the last exercise together.

# Exercise 3: The Telco customer churn

This exercise is based on a file downloaded from the IBM website (it's probably synthetic data, but it is still interesting to examine). The original file is available [here](https://www.ibm.com/communities/analytics/watson-analytics-blog/guide-to-sample-datasets/), but you can also load it from our github repo.

```{r read telco}
telco_churn <- read_csv("https://raw.githubusercontent.com/adisarid/Riskified_training/master/datasets/WA_Fn-UseC_-Telco-Customer-Churn.csv")

glimpse(telco_churn)
```

**Question 1:** Using a scatter plot (`geom_point`), examine the distribution of TotalCharges as a function of tenure.

   1. What can you deduce from the way the points are scattered?
   2. Use facets to split the chart by contract, and add smoothing (`stat_smooth`). What does this tells you?

**Question 2:** Examine the distribution of monthly charges versus churn.

   1. Is there a relationship between the monthly payments and the churn?
   2. Now, add a chart which examines the monthly charges versus churtn but also separates the the internet service type. What do you deduce now that you see the new chart?
   
**Question 3:** Build a chart that shows the number of churning and remaining clients versus the type of services they received. Are you able to explain the phenomena in the last part better?


**Checkpoint 4: If you reached here you have completed exercise 3. A-M-A-Z-I-N-G!.**

   a. Stick your green sticky note. 
   b. You can take a break and/or continue with the exercises. 
   c. When I see enough green sticky notes around we will solve the last exercise together.

# Exercise 4: Chart branding - themes and scales

In this short exercise you will have to control "themes" and "scales" to generate a Riskified branded charts. We will also add a logo to the charts.

**Chart 1:** We start by a chart we have already seen (miles per galon as afunction of displacement, with gear and horse power).

First, for the logo you must install the `rsvg` packages (`install.packages("rsvg")`). Then, you only need to replace the "???" in the proper places. Note that the final command is `grid.raster` from package `grid` (which you may also need to install). This puts the logo into the chart.

```
#install.packages("rsvg")
riskified_logo <- rsvg::rsvg("https://3syq88kbbdj40jbkf2ssy21e-wpengine.netdna-ssl.com/wp-content/themes/riskified/images/logo.svg")

ggplot(mtcars, aes(x = disp, y = mpg)) + 
   geom_point(aes(size = hp, color = factor(gear))) + 
   scale_???_???(values = c(???, ???, ???)) +
   ggtitle("???") + 
   labs(caption = "???") + 
   theme_??()

grid::grid.raster(riskified_logo, x = 0.07, y = 0.03, just = c("left", "bottom"), width = unit(0.7, 'inches'))
```

**Chart 2:** Now, we will do the same but to a different chart.

```
ggplot(telco_churn, aes(x = InternetService, fill=Churn)) + 
   geom_bar(position = "stack", color = ???) + 
   scale_???_???(???) + 
   ggtitle("Comparing Telco's churn over internet service type") +
   theme_??()
```


# Further study

* Generating maps
* Generating interactive charts:
   * Why and when should you do that?
   * Generating interactive charts using `plotly`
   * Generating interactive charts using `ggvis`, and describe the main differences between ggvis and ggplot2
   * Interactive maps using leaflet
* Generating animated charts using `gganimate`